---
title: "Homework2"
output:
  pdf_document:
    latex_engine: xelatex
---


```{r}
options(repos = c(CRAN = "https://cran.rstudio.com"))

install.packages(c(
  "tidyverse",
  "ggplot2",
  "earth",
  "mgcv",
  "pdp",
  "patchwork",
  "caret"
))
```



## Point 1. Read in the data and partition into training and test sets


```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Libraries to be used in this:
```{r}
library(tidyverse)
library(ggplot2)
library(earth)
library(mgcv)
library(pdp)
library(patchwork)
library(caret)
```

reading in data:
```{r}
setwd("/Users/karolinawiesiolek/Desktop/Mailman Spring 2026/DSII/HW2")

college <- read.csv("College.csv", stringsAsFactors = FALSE)

dim(college)
names(college)
head(college)
summary(college)
```

rename first column:
```{r}
if (names(college)[1] != "College") {
  names(college)[1] <- "College"
}

str(college)
```

data partition:
```{r}
set.seed(2026)

n <- nrow(college)
train_id <- sample(seq_len(n), size = round(0.8 * n))

train <- college[train_id, ]
test  <- college[-train_id, ]

dim(train)
dim(test)
```

Defining RMSE function:
```{r}
rmse <- function(pred, truth) {
  sqrt(mean((pred - truth)^2))
}
```

## Point 2. Part (a): Smoothing spline using Top10perc only

#2.1 Fitting smoothing splines across a range of degrees of freedom
```{r}
df_grid <- 3:12

spline_fits <- lapply(df_grid, function(d) {
  smooth.spline(
    x = train$Top10perc,
    y = train$Outstate,
    df = d,
    cv = TRUE
  )
})

cv_table <- data.frame(
  df = df_grid,
  cv_score = sapply(spline_fits, function(fit) fit$cv.crit)
)

cv_table
```

# 2.2 Choosing the best degree of freedom
```{r}
best_df <- cv_table$df[which.min(cv_table$cv_score)]
best_df
```


#2.3 Creating data for plotting fitted curves
```{r}
x_grid <- seq(min(train$Top10perc), max(train$Top10perc), length.out = 200)

plot_data_list <- lapply(seq_along(df_grid), function(i) {
  pred_i <- predict(spline_fits[[i]], x = x_grid)
  
  data.frame(
    Top10perc = pred_i$x,
    Outstate_hat = pred_i$y,
    df = paste("df =", df_grid[i])
  )
})

plot_data <- bind_rows(plot_data_list)

```

#2.4 Plotting fitted curves for each degree of freedom
```{r}
ggplot(train, aes(x = Top10perc, y = Outstate)) +
  geom_point(alpha = 0.5, color = "gray40") +
  geom_line(
    data = plot_data,
    aes(x = Top10perc, y = Outstate_hat),
    color = "blue",
    linewidth = 1
  ) +
  facet_wrap(~ df, ncol = 2) +
  labs(
    title = "Smoothing spline fits across degrees of freedom",
    x = "Top10perc",
    y = "Outstate"
  ) +
  theme_minimal()
```

#2.5 Plotxting the optimal smoothing spline fit
```{r}
best_spline <- smooth.spline(
  x = train$Top10perc,
  y = train$Outstate,
  df = best_df,
  cv = TRUE
)

best_pred <- predict(best_spline, x = x_grid)

best_plot_df <- data.frame(
  Top10perc = best_pred$x,
  Outstate_hat = best_pred$y
)
```

```{r}
ggplot(train, aes(x = Top10perc, y = Outstate)) +
  geom_point(alpha = 0.5, color = "gray40") +
  geom_line(
    data = best_plot_df,
    aes(x = Top10perc, y = Outstate_hat),
    color = "red",
    linewidth = 1.2
  ) +
  labs(
    title = paste("Optimal smoothing spline fit (df =", best_df, ")"),
    subtitle = "Selected using the smallest CV score",
    x = "Top10perc",
    y = "Outstate"
  ) +
  theme_minimal()
```

#2.6  test error computing for the selected spline model
```{r}
spline_test_pred <- predict(best_spline, x = test$Top10perc)$y
spline_test_rmse <- rmse(spline_test_pred, test$Outstate)

spline_test_rmse
```

#2.7 answer for Point 2a
At lower degrees of freedom, the smoothing spline is very smooth and may underfit the relationship between Top10perc and Outstate. As the degrees of freedom go up, the spline becomes more flexible and can capture more curvature in the data. When the degrees of freedom become too large, the fitted curve may become too wiggly and begin fitting noise rather than the underlying trend.

I selected the optimal degree of freedom using the cross-validation criterion returned by smooth.spline(..., cv = TRUE). I chose the degree of freedom with the smallest CV score.

## Point 3. Part (b): Multivariate Adaptive Regression Splines (MARS)
#3.1 data prepp for MARS 
```{r}
train_mars <- train %>% select(-College)
test_mars  <- test %>% select(-College)

str(train_mars)
```

#3.2  MARS model fitting
```{r}
set.seed(2026)

mars_fit <- earth(
  Outstate ~ .,
  data = train_mars,
  degree = 2,
  trace = 0
)

summary(mars_fit)
```

#3.3 Reportting the regression function
```{r}
formula(mars_fit)
coef(mars_fit)
```

#3.4  variable importance
```{r}
evimp(mars_fit)
```

#3.5 Partial dependence plot for an arbitrary predictor
```{r}
mars_pdp <- partial(
  object = mars_fit,
  pred.var = "Expend",
  train = train_mars
)

head(mars_pdp)
```

```{r}
autoplot(mars_pdp, contour = FALSE) +
  labs(
    title = "Partial dependence plot for Expend in the MARS model",
    x = "Expend",
    y = "Predicted Outstate"
  ) +
  theme_minimal()
```


#3.6 test error for the MARS model
```{r}
mars_test_pred <- predict(mars_fit, newdata = test_mars)
mars_test_rmse <- rmse(mars_test_pred, test_mars$Outstate)

mars_test_rmse
```

#3.7 answer for part 3b

The MARS model fits many small lines instead of one straight line. This helps it capture curved (nonlinear) relationships and interactions between variables automatically.The partial dependence plot for Expend shows how predicted out-of-state tuition changes as instructional spending changes, while the other variables are kept roughly the same.

## Point 4. Part (c): Generalized Additive Model (GAM)

#4.1 data for GAM
```{r}
train_gam <- train %>% select(-College)
test_gam  <- test %>% select(-College)
```

#4.2 fitting GAM model
```{r}
set.seed(2026)

gam_fit <- gam(
  Outstate ~ 
    s(Apps, k = 10) +
    s(Accept, k = 10) +
    s(Enroll, k = 10) +
    s(Top10perc, k = 10) +
    s(Top25perc, k = 10) +
    s(F.Undergrad, k = 10) +
    s(P.Undergrad, k = 10) +
    s(Room.Board, k = 10) +
    s(Books, k = 10) +
    s(Personal, k = 10) +
    s(PhD, k = 10) +
    s(Terminal, k = 10) +
    s(S.F.Ratio, k = 10) +
    s(perc.alumni, k = 10) +
    s(Expend, k = 10) +
    s(Grad.Rate, k = 10),
  data = train_gam,
  method = "REML",
  select = TRUE
)

summary(gam_fit)
```


#4.3 nonlinear terms plot
```{r}
plot(gam_fit, pages = 1, shade = TRUE, residuals = TRUE)
```

#4.4 test error for the GAM model
```{r}
gam_test_pred <- predict(gam_fit, newdata = test_gam)
gam_test_rmse <- rmse(gam_test_pred, test_gam$Outstate)

gam_test_rmse
```

#4.5 Written answer for part (c)

The GAM model lets each predictor have its own smooth, possibly curved relationship with Outstate (tuition). If the degrees of freedom are greater than 1, it means the relationship is nonlinear (curved). The plots show these relationships. Some predictors look linear, while others show curves or changing effects.The test RMSE tells us how well the model predicts tuition on the test data (prediction accuracy).


## Point 5. Part (d): Would I favor a MARS model over a linear model in this dataset?

#5.1 linear model for comparison
```{r}
train_lm <- train %>% select(-College)
test_lm  <- test %>% select(-College)

lm_fit <- lm(Outstate ~ ., data = train_lm)

summary(lm_fit)
```

#5.2 test error for the linear model
```{r}
lm_test_pred <- predict(lm_fit, newdata = test_lm)
lm_test_rmse <- rmse(lm_test_pred, test_lm$Outstate)

lm_test_rmse
```

#5.3 predictive performance across models
```{r}
results <- data.frame(
  Model = c(
    "Smoothing spline (Top10perc only)",
    "MARS",
    "GAM",
    "Linear model"
  ),
  Test_RMSE = c(
    spline_test_rmse,
    mars_test_rmse,
    gam_test_rmse,
    lm_test_rmse
  )
)

results[order(results$Test_RMSE), ]
```


#5.4  answer for part (d)

In this dataset, I would pick the MARS model over a linear model if the MARS model achieves a clearly lower test RMSE. A lower test RMSE would suggest that nonlinear relationships and interactions are important for predicting out-of-state tuition, and that a simple linear model may be too restrictive. Because tuition is likely influenced by multiple institutional reasons in complex ways, MARS can be useful since it automatically captures nonlinearities and interactions. However, if the MARS and linear model have very similar test errors, I would prefer the linear model because it is simpler and easier to interpret.


## Point 6. Part (e): In practical applications, is MARS superior to a linear model?
6.1  answer for part (e)

In practical applications, I would not say that a MARS model is always superior to a linear model. The better choice depends on the purpose of the analysis and the structure of the data.

A MARS model is often prefered when the main goal is prediction and the data contain nonlinear relationships or interactions that are difficult to specify ahead of time. Because it is more flexible, it can produce better predictive accuracy when the true relationships are not linear.

A linear model is often preferable when ability to interpret, simplicity, and ease of communication are more important. Linear models are easier to explain, easier to report, and less complex. If the linear model performs nearly as well as the MARS model, the linear model may be the better practical choice.

Therefore, in practice, I would compare both models on test-set performance and then decide whether the predictive gain from MARS is large enough to justify the additional complexity.


##Point 7. Final comparison table
```{r}
results
```

















